---
title: "Reproducible research" 
subtitle: "collaboration and digital data" 
author: "Alyanne de Haan, Mark Teunis"
output: 
  ioslides_presentation:
     widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## prepare

```{r, echo = FALSE, out.width=780, fig.cap=""}
knitr::include_graphics(
  here::here(
    "images",
    "twitter_branch.png"
    )
)
```

Projects are changing all the time.

## reproducible workflows

https://www.youtube.com/watch?v=s3JldKoA0zw

## Does it take a lot of time?

```{r, echo = FALSE, out.width=700, fig.cap=""}
knitr::include_graphics(
  here::here(
    "images",
    "time.png"
    )
)
```

sheet from [Olivier Gimenez](https://oliviergimenez.github.io/reproducible-science-workshop/), shared Creative Commons Attribution [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). 

## Split: sharing data

regular:

 - I don’t use code (R, Python, Matlab, Julia, etc) to analyse my data
 - I would like to use a tidy data format to make my data machine readable
 - I would like to use onedrive / google docs / similar software for version controlling my work.

next level:

 - I use code to analyse my data and already know all about tidy data.
 - I would like to automate my communication of workflows
 - I would like to use github for version controlling my work.

## Exercises regular

 - B1: evaluate your emailing of data / study information. 
 - B2: understanding possible issues when sharing excel files

## Exercises next level

 - B1: evaluate your emailing of data / study information. 
 - C1 + C2 + C3 : get started with github

## Location

Breakout rooms, groups of 5

back here in 30 minutes


