---
title: "Introduction"
author: "[Marc Teunis, PhD](https://orcid.org/0000-0002-3496-6669); [Alyanne de Haan, PhD](https://orcid.org/0000-0003-0896-0906)"
affiliation: "HU University of Applied Sciences"
date: "`r Sys.time()`"
output:
  xaringan::moon_reader:
    includes:
      after_body: 'insert-logo.html'
    nature:
      ratio: '16:9'
subtitle: "Reproducible Open Science"
always_allow_html: yes
widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

image_dir <- here::here(
  "images"
)

## packages
library(tidyverse)
library(nlme)
library(DiagrammeR)
```

# Data, methods and logic

[*Brown, Kaiser & Allison, PNAS, 2018*](https://doi.org/10.1073/pnas.1708279115)

"...in science, three things matter:

--
 ###- the data,
 ###- the methods used to collect the data [...], and
 ###- the logic connecting the data and methods to conclusions,
--


 ### everything else is a distraction."
---

# `Gollums` lurking about

```{r, dpi = 120}
knitr::include_graphics(
  file.path(
    image_dir,
    "gollum_climbing.jpg"
  )
)
```


--
"In one case, a group accidentally used reverse-coded variables, making their conclusions the opposite of what the data supported."


--
"In another case, authors received an incomplete dataset because entire categories of data were missed; when corrected, the qualitative conclusions did not change, but the quantitative conclusions changed by a factor of \>7"


--
[Brown, Kaiser & Allison, 2018; PNAS](https://doi.org/10.1073/pnas.1708279115)
---

# Why do we need Reproducible (Open) Science?

 ###- To assess validity of science and methods we need access to data, methods and conclusions
 ###- To learn from choices other researchers made
 ###- To learn from omissions, mistakes or errors
 ###- To prevent publication bias (also negative results will be available in reproducible research)
 ###- To be able to re-use and/or synthesize data (from many and diverse sources)

[Nature Collection on this topic](https://www.nature.com/collections/prbfkwmwvz)
---

# The *GUI problem*

How would you 'describe' the steps of an analysis or creation of a graph when you use GUI\* based software?

```{r, dpi = 80}
knitr::include_graphics(
  file.path(
    "images",
    "messy_steps.jpg"
  )
)
```

--
***"You can only do this using code, so it is (basically) impossible in a GUI"***

<p style="font-size:14px">
-- 
\*[Graphical User Interface (GUI)...is a form of user interface that allows users to interact with electronic devices through graphical icons and audio indicator such as primary notation, instead of text-based user interfaces, typed command labels or text navigation...](https://en.wikipedia.org/wiki/Graphical_user_interface)

</p>
---

# Programming is essential for Reproducible (Open) Science

 ###- Only programming an analysis (or creation of a graph) records every step
 ###- The script(s) function as a (data) analysis journal
 ###- Code is the logic that connects the data and methods to conclusions
 ###- Learning to use a programming language takes time but pays of at the long run (for all of science)
 ###- And it is fun!
 ###- But, it is not easy to learn and will cost time
---

# To replicate a scientific study we need at least:

####> - Scientific context, research questions and state of the art [P]
####> - (Experimental) model or characteristics of population or matter studied [P]
> - Data that was generated and corresponding meta data [D, C]
> - **Exact** (experimental) design of the study [P, D, C]
> - Exploratory data analysis of the data [P, C]
> - **Exact** methods that were used to conduct any formal inference [P, C]
> - Model diagnostics [*C*]
> - Interpretations of the (statistical) model results/model fitting process [*P*, *C*]
> - Conclusions and academic scoping of the results [<mark>P</mark>, *C*]
> - **Access to all of the above** [<mark>OAcc, OSrc</mark>]

P = Publication, D = Data, C = Code, OAcc = Open Access, OSrc = Open Source
--

### **= FAIR**


---

# FAIR Data

The FAIR principles:

 ###- **F** = *Findable*
 ###- **A** = *Accessible*
 ###- **I** = *Interoperable*
 ###- **R** = *Re-usable*

GO FAIR - Available from: [https://www.go-fair.org/go-fair-initiative/](https://www.go-fair.org/go-fair-initiative/)
---

# **F**indable data

 ###- Datasets should be described, identified and registered or indexed in a clear and unequivocal manner
 ###- Datasets should be findable by machines **and** humans

Examples:

 ###- [Digital Object Identifier - DOI](http://n2t.net/e/ark_ids.html)
 ###- [Archival Resource Key - ARK](https://www.doi.org/)
 ###- [Handle - HNDL](https://www.handle.net/)
 ###- [Persistent Uniform Resource Locator - PURL](https://sites.google.com/site/persistenturls/)
---

# **A**ccesible data

 ###- The (meta)data should be freely available
 ###- Does not mean access to the data is free
 ###- Accessibility is communication and linkage
 ###- Condition under which the associated data are accessible should be open and free.
 ###- Heavily protected and private data can be FAIR data. But is not Open Science
 ###- Licence needs to be connected to the metadata
---

# **I**nteroperable data

 ###- To enhance connecting and enriching data
 ###- Information about analysis, platform, technology, software version, entity ids
 ###- Controlled vocabularies and ontologies
 ###- (Meta)data use a formal, accessible, shared, and broadly applicable language for knowledge 
 ###- representation. (metadata standards)
 ###- (Meta)data use vocabularies that follow FAIR principles
 ###- (Meta)data include qualified references to other (meta)data (resolvers)
---

# **R**e-usable data

 ###- (Meta)data are released with a clear and accessible data usage license
 ###- (Meta)data are associated with detailed provenance
 ###- (Meta)data meet domain-relevant community standards
 ###- Data should be stored in a non-proprietary format (e.g. not `.xls(x)` or `.sav`)
---

# Workflow 'Mermaid' - Current

```{r, echo=TRUE}
library(DiagrammeR)
mermaid("
graph LR
 CellCultureExperiment-->LabJournal
 MicrobiologyExperiment-->LabJournal
 LabJournal-->XLSX
 XLSX-->ExcelOnLaptop
")

```
---

# Workflow 'Mermaid' - Future

```{r, echo=TRUE}
library(DiagrammeR)
mermaid("
graph LR
 CellCultureExperiment-->Robot
 MicrobiologyExperiment-->Robot
 Robot-->ElectronicLabJournal
 ElectronicLabJournal-->Dashboard
")

```
---

# Ontox workflow

[mermaid markdown](https://github.com/ontox-hu/workflows/blob/main/WorkflowONTOX.md)

