---
title: "Miscellaneous"
author: "[Marc Teunis, PhD](https://orcid.org/0000-0002-3496-6669); [Alyanne de Haan, PhD](https://orcid.org/0000-0003-0896-0906)"
date: "`r Sys.time()`"
output:
  ioslides_presentation:
    logo: cropped-logo-ontox_orange-01-01-2.webp
  beamer_presentation: default
  powerpoint_presentation: default
subtitle: Reproducible Research & FAIR Data
always_allow_html: yes
widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

image_dir <- here::here(
  "images"
)

## packages
library(tidyverse)
library(nlme)
library(DiagrammeR)
```

## Version controlled findable data

Databases that provide access to the data usually work with time-stamped identifiers for updates

```{r, dpi=50}
knitr::include_graphics(
  here::here(
    "images",
    "pubchem_version.png"
  )
)
```

## FAIR data point (FDP)

-   A formal API that hosts metadata of datasets and that can be *visited* by computers (scripts) and humans
-   To empower people to store their data as FAIR data on the web
-   Consists of ["FAIR Data Point 'API' specification"](https://specs.fairdatapoint.org/); Reference implementation - authentication system and a web client that hoists the FDP

Example: [VODAN](https://www.vodan-totafrica.info/about-vodan)

<p style="font-size:14px">

<https://github.com/FAIRDataTeam/FAIRDataPoint>

</p>

<p style="font-size:14px">

<https://specs.fairdatapoint.org/>

</p>

## Metadata standards

To enable harmonization and inter-operability standards have been defined

Examples:

-   [DCAT](https://www.w3.org/TR/vocab-dcat-2/)
-   [Dublin Core](https://www.dublincore.org/)
-   [Linked Data Platform](https://www.w3.org/TR/ldp/)

## Dublin Core [DC]

Namespace: terms to include in the metadata schema

A namespace has properties:

*abstract, accessRights, accrualMethod, accrualPeriodicity,* *accrualPolicy, alternative, audience, available, bibliographicCitation,* *conformsTo, contributor, coverage, created, creator, date, ...*

[DC terms namespace](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/)

DC has definitions for RDF and XML.

**When we look at the different namespaces, we discover that the DC is actually an ontology ;-)**

# Interoperability example

| attribute  | property                                                   |
|-----------------|-------------------------------------------------------|
| id:        | <GO:0000016>                                               |
| name:      | lactase activity                                           |
| namespace: | molecular_function                                         |
| def:       | "Catalysis of the reaction: lactose + H2O = D-glucose + D-galactose." EC:3.2.1.108] |
| synonym:   | "lactose galactohydrolase activity" EXACT [EC:3.2.1.108]   |
| ...        | ...                                                        |

<p style="font-size:14px">

[Gene Ontology Resource](http://geneontology.org/)

</p>
---

# GENE annotation with this GO term:

| attribute        | property                                                       |
|------------------|----------------------------------------------------------------|
| Submitted name   | Lactase, gene 1                                                |
| Gene             | lct.1                                                          |
| Organism         | Xenopus tropicalis (Western clawed frog) (Silurana tropicalis) |
| Status           | Unreviewed                                                     |
| Annotation score | Protein predicted 

---


# Licences

 ###- CC ZERO (No limits)
 ###- CC BY (Attribution)
 ###- CC BY-SA (Attribution + Remix/Reuse/Extend)
 ###- CC BY-ND (Attribution + NoDerivatives - reuse as is)
 ###- CC BY-NC (Attribution + NonCommercial)
 ###- CC BY-NC-SA (Combination)
 ###- CC BY-NC-ND (Restrictive combination)
 ###- Copyright
---

## Workflows - Current

```{r}
knitr::include_graphics(
  here::here(
    "images",
    "state-of-the-art-data-collection.jpg"
  )  
)
```

<p style="font-size:14px">

Solle, D. Be FAIR to your data. Anal Bioanal Chem 412, 3961--3965 (2020). <https://doi.org/10.1007/s00216-020-02526-7>

</p>
---

# Workflows - Future

```{r}
knitr::include_graphics(
  here::here(
    "images",
    "future-data-collection.jpg"
    )
  )
```
<p style="font-size:14px">

Solle, D. Be FAIR to your data. Anal Bioanal Chem 412, 3961--3965 (2020). <https://doi.org/10.1007/s00216-020-02526-7>

</p>
---

# Contents

 ###- Reproducible (Open) Science
 ###- FAIR Data
 ###- Work Flows

The complete source code for the webinars and all dependent data, and files can be found on [Github.com/ontox-hu](https://github.com/ontox-hu/data_management_workshop).
---

# Introducing Reproducible (Open) Science

 ###1. When things go wrong
 ###2. Why Reproducible (Open) Science?
 ###3. The need for learning to program
 ###4. FAIR data

### Reproducible (Open) Science = Reproducible Research + Open Science
---

# Metadata

 ###- To find data, datasets should be accompanied by rich metadata ("data about the data")
 ###- Metadata can adhere to a standard, but not necessarily
 ###- At least describe: 
 
  - title
  - origin
  - date 
  - variable description
  - owner 
  - licence
  - version
  - format
---
