---
title: "Reproducible Research Example"
author: '[Marc Teunis, PhD](https://orcid.org/0000-0002-3496-6669); [Alyanne de Haan, PhD](https://orcid.org/0000-0003-0896-0906)'
date: '`r Sys.time()`'
output:
  ioslides_presentation:
    highlight: tango
    logo: "cropped-logo-ontox_orange-01-01-2.webp"
subtitle: Reproducible Research
always_allow_html: yes
widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

image_dir <- here::here(
  "images"
)

## packages
library(tidyverse)
library(nlme)
```


## A short example of Reproducible (Open) Science

Assume we have the following question:
"Which of 4 types of chairs takes the least effort to arise from when seated in?"
We have the following setup:

 - 4 different types of chairs
 - 9 different subjects (probably somewhat aged)
 - Each subject is required to provide a score (from 6 to 20, 6 being very lightly strenuous, 20 being extremely strenuous) when arising from each of the 4 chairs. There is some 'wash-out' time in between the trials. The chair order is randomised.

To analyze this experiment statistically, the model would need to include: the rating score as the **measured (or dependent) variable**, the type of chair as the **experimental factor** and the subject as the **blocking factor**

## Mixed effects models

A typical analysis method for this type of randomized block design is a so-called 'multi-level' or also called 'mixed-effects' or 'hierarchical' model. An analysis method much used in clinical or biological scientific practice. 

## What do we minimally need, to replicate the science of this experiment? {.build}

 >- the data 
 >- an exploratory graph 
 >- a statistical model 
 >- the statistical model results
 >- a model diagnostic
 >- some conclusions 
 
In the next few slides, I will hopefully convince you of the power of (literate) programming to communicate such an analysis. 

<p style="font-size:14px">[Example reproduced from: Pinheiro and Bates, 2000, _Mixed-Effects Models in S and S-PLUS_, Springer, New York.](https://cran.r-project.org/web/packages/nlme/index.html)</p>
 
## The data of the experiment

<p style="font-size:14px">[Wretenberg, Arborelius & Lindberg, 1993](https://doi.org/10.1080/00140139308967910)</p>


```{r, echo=TRUE}
library(nlme)
ergoStool %>% as_tibble()
```

## An exploratory graph
```{r}
set.seed(123)
plot_ergo <- ergoStool %>%
  ggplot(aes(x = reorder(Type, effort), y = effort)) + 
  geom_boxplot(colour = "darkgreen", outlier.shape = NA) + 
  geom_jitter(aes(colour = reorder(Subject, -effort)), 
              width = 0.2, size = 3) +
  scale_colour_manual(values = c("red","blue", "green", "darkblue", "darkgreen", "purple", "grey", "black", "darkgrey")) +
  ylab("Effort (Borg scale score)") +
  xlab("Chair type") + 
  guides(colour=guide_legend(title="Subject id")) +
  theme_bw()
plot_ergo
```

## Mind the variability per subject, what do you see?

 - Can you say something about within-subject variability (note 'Minster Blue')?
 - Can you say something about between-subject variability (note 'Mister Green', vs 'Mister Black')?
 - Which chair type takes, on average the biggest effort to arise from?
 
```{r, fig.width=5, fig.height=3}
plot_ergo
```

## The statistical questions

 1. Which chair type takes, on average the biggest effort to arise from? (ANOVA / MEM, fixed effects)
 - Do individual (within subject) differences play a role in appointing a average score to a chair type? (MEM, random effects)
 - Does variability between subjects play a role in determining the 'best' chair type (ANOVA / MEM, confidence intervals)

## The statistical model 
Statistical models (in R) can be specified by a `model formula`. The left side of the formula is the dependent variable, the right side are the 'predictors'. Here we include a `fixed` and a `random` term to the model (as is common for mixed-effects models)

```{r, echo=TRUE, eval=FALSE}
library(nlme)
```
```{r, echo=TRUE}
ergo_model <- lme(
  data = ergoStool, # the data to be used for the model
  fixed = effort ~ Type, # the dependent and fixed effects variables
  random = ~1 | Subject # random intercepts for Subject variable
)
```

<p style="font-size:18px">The `lme()` function is part of the [`{nlme}`](https://cran.r-project.org/web/packages/nlme/index.html) package for mixed effects modelling in R</p>

<p style="font-size:18px">Example reproduced from: [Pinheiro and Bates, 2000, _Mixed-Effects Models in S and S-PLUS_, Springer, New York.](https://cran.r-project.org/web/packages/nlme/index.html)</p>

## The statistical results
```{r}
result <- ergo_model %>% summary() 
result$tTable %>% as.data.frame() %>% knitr::kable()
```

## Model diagnostics {.build}

 >- Diagnostics of a fitted model is the most important step in a statistical analysis
 >- In most scientific papers the details are lacking 
 >- Did the authors omit to perform this step? Or did they not report it?
 >- If you do not want to include it in your paper, put it in an appendix!
 
A residual plot shows the 'residual' error ('unexplained variance') after fitting the model. Under the Normality assumption standardized residuals should:
 
 >1. Be normally distributed around 0
 >1. Display no obvious 'patterns'
 >1. Should display overall equal 'spread' above and below 0 ('assumption of equal variance')
 
## Residual plot
```{r, echo=TRUE}
plot(ergo_model) ## type = 'pearson' (standardized residuals)
```

## The conclusions in a plot
```{r}
# install.packages("ggsignif")
library(ggsignif)
p_values <- result$tTable %>% as.data.frame()
annotation_df <- data.frame(Type=c("T1", "T2"), 
                            start=c("T1", "T1"), 
                            end=c("T2", "T3"),
                            y=c(16, 14),
                            label=
                              paste("p-value:",
                              c(
                              formatC(
                                p_values$`p-value`[2], digits = 3),
                              formatC(
                                p_values$`p-value`[3], digits = 3)
                              )
                            )
                          )
                            
set.seed(123)
ergoStool %>%
  ggplot(aes(x = reorder(Type, effort), 
             y = effort)) + 
  geom_boxplot(colour = "darkgreen", 
               outlier.shape = NA) + 
  geom_jitter(aes(
    colour = reorder(Subject, -effort)), 
    width = 0.2, 
    size = 3) +
  scale_colour_manual(
    values = c(
      "red", "blue","green", 
      "darkblue", "darkgreen", 
      "purple", "grey", "black", 
      "darkgrey")) +
  ylab("Effort (Borg scale score)") +
  xlab("Chair type") + 
  guides(colour=guide_legend(title="Subject id")) +
  ylim(c(6,20)) +
  geom_signif(
    data=annotation_df,
    aes(xmin=start, 
    xmax=end, 
    annotations=label, 
    y_position=y),
    textsize = 5, vjust = -0.2,
    manual=TRUE) +
  theme_bw() -> plot_ergo
plot_ergo
```

